---
title: "Exploring the Most Popular Content on Netflix"
author: "Hymie Israel"
format:
  html:
    toc: true
    toc-depth: 2
    toc-expand: false
    code-fold: true
---
We'll be analyzing Netflix's Global Top 10 viewing data, which provides fascinating insights into worldwide streaming preferences and trends. This dataset captures weekly performance metrics for the most popular content on Netflix's platform.

##### Setup Code
```{r setup, include=FALSE}
# LOAD & INSTALL REQUIRED PACKACGES
if(!require("tidyverse")) install.packages("tidyverse")
library(tidyverse)
library(readr)
library(dplyr)

if(!require("DT")) install.packages("DT")
library(DT)

if(!require("knitr")) install.packages("knitr")
library(knitr)

if(!require("scales")) install.packages("scales")
library(scales)
if(!require("ggrepel")) install.packages("ggrepel")
library(ggrepel)

if(!require("rnaturalearth")) install.packages("rnaturalearth")
if(!require("rnaturalearthdata")) install.packages("rnaturalearthdata")
library(rnaturalearth)
library(rnaturalearthdata)

if(!require("sf")) install.packages("sf")
library(sf)


# CREATE DATA DIRECTORY IF IT DOESN'T EXIST
if(!dir.exists(file.path("data", "mp01"))){
    dir.create(file.path("data", "mp01"), showWarnings=FALSE, recursive=TRUE)
}
# DOWNLOAD DATA IF IT DOESN'T EXIST LOCALLY
GLOBAL_TOP_10_FILENAME <- file.path("data", "mp01", "global_top10_alltime.csv")

if(!file.exists(GLOBAL_TOP_10_FILENAME)){
    download.file("https://www.netflix.com/tudum/top10/data/all-weeks-global.tsv", 
                  destfile=GLOBAL_TOP_10_FILENAME)
}

COUNTRY_TOP_10_FILENAME <- file.path("data", "mp01", "country_top10_alltime.csv")

if(!file.exists(COUNTRY_TOP_10_FILENAME)){
    download.file("https://www.netflix.com/tudum/top10/data/all-weeks-countries.tsv", 
                  destfile=COUNTRY_TOP_10_FILENAME)
}

```

```{r ref.label='setup', eval=FALSE, echo=TRUE}

```

```{r data cleaning, include=FALSE}
# LOAD & CLEAN GLOBAL DATA
GLOBAL_TOP_10 <- read_tsv(GLOBAL_TOP_10_FILENAME)

str(GLOBAL_TOP_10)
glimpse(GLOBAL_TOP_10)

GLOBAL_TOP_10 <- GLOBAL_TOP_10 |> 
  mutate(season_title = if_else(season_title == "N/A", NA_character_, season_title))


# LOAD & CLEAN COUNTRY DATA
COUNTRY_TOP_10 <- read_tsv(COUNTRY_TOP_10_FILENAME, na = "N/A")

str(COUNTRY_TOP_10)
glimpse(COUNTRY_TOP_10)


```

``` {r ref.label='data cleaning', eval=FALSE, echo=TRUE}

```

```{r initial data exploration, include=FALSE}
# INITIAL EXPLORATION OF DATA
GLOBAL_TOP_10 |> 
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE), caption="Table 1: Global Top 10 of All Time")

format_titles <- function(df){
    colnames(df) <- str_replace_all(colnames(df), "_", " ") |> str_to_title()
    df
}

GLOBAL_TOP_10 |> 
    format_titles() |>
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE),caption=) |>
    formatRound(c('Weekly Hours Viewed', 'Weekly Views'))

GLOBAL_TOP_10 |> 
    select(-season_title) |>
    format_titles() |>
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE)) |>
    formatRound(c('Weekly Hours Viewed', 'Weekly Views'))

```

```{r ref.label='initial data exploration', eval=FALSE, echo=TRUE}

```



## Dataset Overview

Our data contains 9 key variables that paint a comprehensive picture of Netflix's top-performing content:

##### Temporal & Ranking Information:

Week: The specific date for each weekly measurement period

Weekly Rank: Position in the top 10 (1-10) for that week

##### Content Classification:

Category: Content type (Films English, Films Non-English, TV English, TV Non-English)

Show Title: The main title of the content

Season Title: Specific season information (for TV shows) or NA for films

##### Performance Metrics:

Weekly Hours Viewed: Total hours watched globally that week

Weekly Views: Number of unique viewing instances

Runtime: Average runtime per episode/film (in minutes)

Cumulative Weeks In Top 10: Total weeks the title has appeared in top 10

This dataset offers rich opportunities to understand global entertainment consumption patterns, content performance dynamics, and the evolving landscape of streaming preferences across Netflix's worldwide subscriber base.

## Most Recent Global Top 10 Netflix Films
```{r Global Top 10 Netflix Films of All Time table, echo=FALSE}
# DISPLAY GLOBAL TOP 10 NETFLIX FILMS OF ALL TIME
GLOBAL_TOP_10 |> 
    mutate(`runtime_(minutes)` = round(60 * runtime)) |> 
    select(-season_title, 
           -runtime) |>
    format_titles() |>
    head(n=20) |>
    datatable(options=list(searching=FALSE, info=FALSE), caption= "Most Recent Global Top 10 Netflix Films") |>
    formatRound(c('Weekly Hours Viewed', 'Weekly Views'), 0) |>
  formatStyle(
    'Show Title',
    fontStyle = "italic"
  )
```

```{r ref.label= 'Global Top 10 Netflix Films of All Time table', eval=FALSE, echo=TRUE}
```

## Exploratory Questions

Q1. How many different countries does Netflix operate in? (You can use the viewing history as a proxy for countries in which Netflix operates)

```{r question-1, include=FALSE}
# COUNT DISTINCT COUNTRIES
q1 <- n_distinct(COUNTRY_TOP_10$country_name)

# MAP COUNTRIES WHERE NETFLIX OPERATES
world <- ne_countries(scale = "medium", returnclass = "sf")

netflix_countries <- COUNTRY_TOP_10 |>
    distinct(country_iso2) |>
    mutate(has_netflix = "Yes")

world_netflix <- world |>
    left_join(netflix_countries, by = c("iso_a2" = "country_iso2")) |>
    mutate(has_netflix = ifelse(is.na(has_netflix), "No", has_netflix))

world_netflix_plot <- ggplot(world_netflix) +
    geom_sf(aes(fill = has_netflix), color = "white", size = 0.2) +
    scale_fill_manual(values = c("No" = "grey80", "Yes" = "red")) +
    labs(
        title = "Countries Where Netflix Operates",
        fill = "Netflix Available"
    ) +
    theme_minimal()

```

Netflix operates in **`r q1`** different countries based on the viewing history data.
```{r world-netflix-plot, echo=FALSE}
world_netflix_plot
```

```{r ref.label='question-1', eval=FALSE, echo=TRUE}
```

Q2. Which non-English-language film has spent the most cumulative weeks in the global top 10? How many weeks did it spend?

```{r question-2, include=FALSE}
# FIND NON-ENGLISH FILM WITH MOST CUMULATIVE WEEKS IN GLOBAL TOP 10
q2a <- 
GLOBAL_TOP_10 |> 
filter(category == "Films (Non-English)" ) |>
group_by(show_title) |> 
summarise(cumulative_weeks_in_top_10 = n()) |> 
arrange(desc(cumulative_weeks_in_top_10)) |> 
slice(1) |> 
pull(show_title)

q2b <- 
GLOBAL_TOP_10 |> 
filter(category == "Films (Non-English)" ) |> 
group_by(show_title) |> 
summarise(cumulative_weeks_in_top_10 = n()) |>
arrange(desc(cumulative_weeks_in_top_10)) |> 
slice(1) |> 
pull(cumulative_weeks_in_top_10)
```

The non-English-language film that has spent the most cumulative weeks in the global top 10 is ***`r q2a`***   with a total of **`r q2b`** cumulative weeks.

```{r ref.label='question-2', eval=FALSE, echo=TRUE}

```

Q3. What is the longest film (English or non-English) to have ever appeared in the Netflix global Top 10? How long is it in minutes?

```{r question-3, include=FALSE}
# FIND LONGEST FILM IN GLOBAL TOP 10
q3a <- 
GLOBAL_TOP_10 |> 
filter(category %in% c("Films (English)", "Films (Non-English)")) |> 
arrange(desc(runtime)) |> 
slice(1) |>
pull(show_title)

q3b <- 
round(60 * GLOBAL_TOP_10 |> 
filter(category %in% c("Films (English)", "Films (Non-English)")) |>
arrange(desc(runtime)) |>
slice(1) |>
pull(runtime))
```

The longest film to have ever appeared in the Netflix global Top 10 is ***`r q3a`*** with a runtime of **`r q3b`** minutes.

```{r ref.label='question-3', eval=FALSE, echo=TRUE}
```

Q4. For each of the four categories, what program has the most total hours of global viewership?

```{r question-4, include=FALSE}
# FIND PROGRAM WITH MOST TOTAL HOURS OF GLOBAL VIEWERSHIP IN EACH CATEGORY
q4a <- GLOBAL_TOP_10 |> 
group_by(category, show_title) |> 
summarise(total_hours_viewed = sum(weekly_hours_viewed)) |>
arrange(category, desc(total_hours_viewed)) |>
slice_max(total_hours_viewed)

q4 <- datatable(
  q4a,
  colnames = str_replace_all(colnames(q4a), "_", " ") |> str_to_title(),
  rownames = FALSE,
  options = list(
    pageLength = 4,
    searching= FALSE,       
    scrollX = TRUE          
  )
  , caption = "Programs with the most viewership in each category"
) |>
  formatRound(c('total_hours_viewed'), 0) |>
  formatStyle(
    'show_title',
    fontStyle = "italic"
  )
```

The programs with the most total hours of global viewership in each category are: ***`r q4`***

```{r ref.label='question-4', eval=FALSE, echo=TRUE}

```

Q5. Which TV show had the longest run in a country’s Top 10? How long was this run and in what country did it occur?

```{r question-5, include=FALSE}
# FIND TV SHOW WITH LONGEST RUN IN A COUNTRY'S TOP 10
q5a <-
COUNTRY_TOP_10 |>
filter(category == "TV") |>
arrange(desc(cumulative_weeks_in_top_10)) |>
slice(1) |> 
pull(show_title)

q5b <-
COUNTRY_TOP_10 |>
filter(category == "TV") |> 
arrange(desc(cumulative_weeks_in_top_10)) |>
slice(1) |>
pull(cumulative_weeks_in_top_10)

q5c <-
COUNTRY_TOP_10 |> 
filter(category == "TV") |>
arrange(desc(cumulative_weeks_in_top_10)) |>
slice(1) |>
pull(country_name)
```

The TV show with the longest run in a country's Top 10 is ***`r q5a`*** with a run of **`r q5b`** weeks in **`r q5c`**.

```{r ref.label='question-5', eval=FALSE, echo=TRUE}

```

Q6. Netflix provides over 200 weeks of service history for all but one country in our data set. Which country is this and when did Netflix cease operations in that country?

```{r question-6, include=FALSE}
# FIND COUNTRY WHERE NETFLIX CEASED OPERATIONS
q6a <-
COUNTRY_TOP_10 |>
group_by(country_name)|>
summarise(week= n_distinct(week), max(week)) |>
filter(week < max(week)) |>
pull(country_name)

q6b <-
COUNTRY_TOP_10 |> 
group_by(country_name)|>
summarise(week= max(week)) |>
filter(week < max(week)) |>
pull(week) |>
format("%B %d, %Y")
```

Netflix ceased operations in **`r q6a`** on **`r q6b`**.

```{r ref.label='question-6', eval=FALSE, echo=TRUE}

```

Q7. What is the total viewership of the TV show Squid Game? Note that there are three seasons total and we are looking for the total number of hours watched across all seasons.

```{r question-7, include=FALSE}
# CALCULATE TOTAL VIEWERSHIP OF SQUID GAME
q7 <-
GLOBAL_TOP_10 |>
filter(show_title=="Squid Game") |>
summarise(weekly_hours_viewed =sum(weekly_hours_viewed)) |>
pull(weekly_hours_viewed) |>
comma()
```

The total viewership of the TV show *Squid Game* across all seasons is **`r q7`** hours.

```{r ref.label='question-7', eval=FALSE, echo=TRUE}

```

Q8. The movie Red Notice has a runtime of 1 hour and 58 minutes. Approximately how many views did it receive in 2021?

```{r question-8, include=FALSE}
# CALCULATE APPROXIMATE NUMBER OF VIEWS FOR RED NOTICE IN 2021
q8 <-
GLOBAL_TOP_10 |>
filter(show_title=="Red Notice", year(week)==2021) |>
summarise(total_views = sum(weekly_hours_viewed/(118/60))) |>
pull(total_views) |>
comma()
```

*Red Notice* received approximately **`r q8`** views in 2021 based on the assumption that each viewer watched the entire movie.

```{r ref.label='question-8', eval=FALSE, echo=TRUE}

```

Q9. How many Films reached Number 1 in the US but did not originally debut there? That is, find films that first appeared on the Top 10 chart at, e.g., Number 4 but then became more popular and eventually hit Number 1? What is the most recent film to pull this off?

```{r question-9, include=FALSE}
# FIND NUMBER OF FILMS THAT REACHED NUMBER 1 IN US BUT DID NOT DEBUT THERE
q9a <-
COUNTRY_TOP_10 |>
filter(country_name =="United States", category =="Films") |>
group_by(show_title) |>
summarise(
  first_week = min(week),
  first_rank = weekly_rank[week == min(week)][1],
  ever_number_1 = any(weekly_rank == 1),
  most_recent_week = max(week),
  .groups = "drop"
) |>
  filter(ever_number_1 == TRUE, first_rank != 1) |>
  nrow()

q9b  <-
COUNTRY_TOP_10 |>
filter(country_name =="United States", category =="Films") |>
group_by(show_title) |>
summarise(
  first_week = min(week),
  first_rank = weekly_rank[week == min(week)][1],
  ever_number_1 = any(weekly_rank == 1),
  most_recent_week = max(week),
  .groups = "drop"
) |>
  filter(ever_number_1 == TRUE, first_rank != 1) |>
  arrange(desc(most_recent_week)) |>
  slice(1) |>
  pull(show_title)
  
```

The number of films that reached Number 1 in the US but did not originally debut there is **`r q9a`** with the most recent film to pull this off being ***`r q9b`***.

```{r ref.label='question-9', eval=FALSE, echo=TRUE}

```

Q10. Which TV show/season hit the top 10 in the most countries in its debut week? In how many countries did it chart?

```{r question-10, include=FALSE}
# FIND TV SHOW/SEASON THAT HIT TOP 10 IN MOST COUNTRIES IN DEBUT WEEK
q10a <-
COUNTRY_TOP_10 |>
filter(category=="TV") |>
group_by(show_title) |>
filter(week==min(week)) |>
summarise(num_countries = n_distinct(country_name)) |>
arrange(desc(num_countries)) |>
slice(1) |>
pull(show_title)

q10b <-
COUNTRY_TOP_10 |>
filter(category=="TV") |>
group_by(show_title) |>
filter(week==min(week)) |>
summarise(num_countries = n_distinct(country_name)) |>
arrange(desc(num_countries)) |>
slice(1) |>
pull(num_countries)
```

The TV show/season that hit the top 10 in the most countries in its debut week is ***`r q10a`*** in **`r q10b`** countries.

```{r ref.label='question-10', eval=FALSE, echo=TRUE}

```

# *Stranger Things*: A Cultural Phenomenon Set to Conclude

```{r stranger-things, include=FALSE}
# ANALYSIS OF STRANGER THINGS VIEWING DATA
total_series_viewing_hours <-
GLOBAL_TOP_10 |>
filter(show_title == "Stranger Things") |>
summarise(total_viewing_hours = sum(weekly_hours_viewed, na.rm = TRUE)) |>
  pull(total_viewing_hours) |>
  comma()

season_4_consecutive_weeks <-
GLOBAL_TOP_10 |>
filter(season_title == "Stranger Things 4") |>
arrange(desc(week)) |>
summarise(max_weeks = max(cumulative_weeks_in_top_10, na.rm = TRUE))

```

Netflix today announced that the fifth and final season of *Stranger Things* will premiere at the end of 2025. With four groundbreaking seasons already released, *Stranger Things* has cemented its place as one of the most successful and culturally significant series in streaming history by demonstrating extraordinary staying power, multinational appeal, and unmatched viewership records. The series has total viewing hours of **`r total_series_viewing_hours`** with its latest season having **`r season_4_consecutive_weeks`** consecutive weeks in the Top 10 globally following its release in May 2022, the highest across all 4 seasons.

```{r ref.label='stranger-things', eval=FALSE, echo=TRUE}

```

Compared to other popular series, Stranger Things has the highest total viewing hours, whilst being in the top 10 for a fewer amount of weeks. Below are two graphs comparing *Stranger Things* and other popular series.

```{r comparison, echo=FALSE}
# PREPARE DATA FOR COMPARISON GRAPHS
comparison_summary <- GLOBAL_TOP_10 |> 
    filter(show_title %in% c("Stranger Things", "Wednesday", "Bridgerton", "Outer Banks", "Money Heist", "You")) |>
    group_by(show_title) |>
    summarise(
        total_hours = sum(weekly_hours_viewed, na.rm = TRUE),
        weeks_in_top10 = max(cumulative_weeks_in_top_10, na.rm = TRUE)
    ) |>
    arrange(desc(total_hours)) 
```

```{r comparison-graphs, echo=FALSE}
# CREATE COMPARISON GRAPHS
total_hours_viewed <- 
ggplot(comparison_summary, aes(x = reorder(show_title, total_hours), y = total_hours, fill = show_title)) +
    geom_col() +
    coord_flip() +
    scale_y_continuous(labels = comma) +
    labs(
        title = "Total Hours Viewed (All Seasons, Global Top 10)",
        x = "Show",
        y = "Total Hours Viewed"
    ) +
    theme_minimal() +
    theme(
      legend.position = "none",
      axis.text.y = element_text(face = "italic"))

hours_vs_weeks <- 
ggplot(comparison_summary, aes(x = weeks_in_top10, y = total_hours, label = show_title, color = show_title)) +
    geom_point(size = 5) +
    geom_text_repel(
      size= 3.5, 
      max.overlaps=Inf,
      aes(label = paste0("italic('", show_title, "')")),
      parse = TRUE) +
    scale_y_continuous(labels= comma) +
    labs(
        title = "Netflix Shows: Total Hours vs. Weeks in Top 10",
        x = "Weeks in Global Top 10",
        y = "Total Hours Viewed",
        color = "Show Title"
    ) +
    theme_minimal() 
```

```{r show-graphs, echo=FALSE}
# DISPLAY COMPARISON GRAPHS
print(total_hours_viewed)
print(hours_vs_weeks)
```

```{r ref.label='comparison-graphs', eval=FALSE, echo=TRUE}

```

# Hindi Hits, Huge Impact: Netflix India Sets Streaming Records

```{r india, include=FALSE}
# ANALYSIS OF INDIA-ONLY NETFLIX HITS
# List of popular non-Hindi shows to exclude
non_hindi_shows <- c("The Good Doctor", "Friends", "Business Proposal")


us_shows <- COUNTRY_TOP_10 |>
    filter(country_name == "United States") |>
    distinct(show_title)


india_only_hits <- COUNTRY_TOP_10 |>
    filter(country_name == "India") |>
    filter(!show_title %in% us_shows$show_title) |>
    filter(!show_title %in% non_hindi_shows)


india_summary <- india_only_hits |>
    group_by(category, show_title) |>
    summarise(
        weeks_in_top_10 = n(),
        first_week = min(week),
        last_week = max(week),
        .groups = "drop"
    ) |>
    arrange(desc(weeks_in_top_10))


peak_hours <- GLOBAL_TOP_10 |>
    group_by(show_title) |>
    summarise(peak_weekly_hours = max(weekly_hours_viewed, na.rm = TRUE), .groups = "drop")


india_subs_estimate <- india_summary |>
    left_join(peak_hours, by = "show_title") |>
    mutate(
        peak_hours_india = peak_weekly_hours * 0.7,    # assume 70% of hours are from India
        estimated_subs = round(peak_hours_india / 5),  # average 5 hrs/week per subscriber
        duration = paste0(as.character(first_week), " to ", as.character(last_week))
    ) |>
    select(category, show_title, weeks_in_top_10, duration, estimated_subs) |>
    arrange(desc(weeks_in_top_10)) |>
    slice_head(n = 10)   

india_sub_estimate_table <- datatable(
  india_subs_estimate,
  colnames = str_replace_all(colnames(india_subs_estimate), "_", " ") |> str_to_title(),
  rownames = FALSE,
  options = list(
    pageLength = 10,
    searching= FALSE,
    autoWidth = TRUE,       
    scrollX = TRUE          
  )
  , caption = "Top 10 India-Only Netflix Hits and Estimated Subscribers"
) |>
  formatRound(c('estimated_subs'), 0) |>
  formatStyle(
    'show_title',
    fontStyle = "italic"
  )

```

Netflix continues to strengthen its presence in India, the world’s most populous country, as its Hindi content resonates strongly with local audiences. A recent analysis of India’s Top 10 Netflix hits shows remarkable engagement and sustained popularity across both television and films.

```{r show-india-table, echo=FALSE}
india_sub_estimate_table
```

```{r ref.label='india', eval=FALSE, echo=TRUE}

```

*The Great Indian Kapil Show* dominated India’s Top 10 for 41 weeks between March 2024 and August 2025, with an estimated 756,000 subscribers engaging with the show weekly. Following closely are *Mismatched*  and *The Railway Men – The Untold Story of Bhopal 1984*, showcasing the audience’s appetite for compelling storytelling rooted in local culture and history.

Other standout TV series include *Khakee: The Bihar Chapter*, *Kota Factory*, *Maamla Legal Hai*, *Rana Naidu*, and *Yeh Kaali Kaali Ankhein*.

On the films side, *Haseen Dillruba* and *Sooryavanshi* demonstrate that Netflix’s cinematic offerings also enjoy strong, sustained viewership across the country.

These Indian hits underline Netflix’s successful strategy of delivering locally relevant content while maintaining global production standards. With engagement metrics pointing to millions of estimated Indian subscribers actively enjoying Hindi programming, the platform is well-positioned for long-term growth in India, further cementing its role as a key player in one of the fastest-growing entertainment markets in the world.

# Netflix Global Blockbusters: Uniting Audiences Across the World

```{r blockbusters, include=FALSE}
# ANALYSIS OF GLOBAL BLOCKBUSTERS
global_hits_summary <- GLOBAL_TOP_10 |>
    group_by(show_title, category) |>
    summarise(
        total_weeks_in_top_10 = n(),
        peak_weekly_hours = max(weekly_hours_viewed, na.rm = TRUE),
        first_week = min(week),
        last_week = max(week),
        .groups = "drop"
    ) |>
    arrange(desc(total_weeks_in_top_10), desc(peak_weekly_hours))


top_global_hits <- global_hits_summary |>
    slice_head(n = 10) |>
    mutate(duration = paste(first_week, "to", last_week)) |>
    select(category, show_title, total_weeks_in_top_10, duration, peak_weekly_hours)

top_global_hits_table <- datatable(
  top_global_hits,
  colnames = str_replace_all(colnames(top_global_hits), "_", " ") |> str_to_title(),
  rownames = FALSE,
  options = list(
    pageLength = 10,
    searching= FALSE,
    autoWidth = TRUE,       
    scrollX = TRUE
  )
  , caption = "Netflix Global Blockbusters: Uniting Audiences Across the World"
) |>
  formatRound(c('peak_weekly_hours'), 0) |>
  formatStyle(
    'show_title',
    fontStyle = "italic"
  )
```

Netflix is celebrating a new era of worldwide entertainment with shows that have captivated audiences across multiple continents. Recent analysis of global Top 10 data reveals the most dominant titles in terms of both sustained engagement and peak weekly hours.

```{r show-top-hits-table, echo=FALSE}
top_global_hits_table
```

```{r ref.label='blockbusters', eval=FALSE, echo=TRUE}

```

*Squid Game* leads the charge, spending 55 weeks in the global Top 10 and achieving a peak weekly viewing of over 571 million hours.

*Stranger Things* maintained an impressive 50 weeks in the Top 10 with 335 million peak hours, solidifying its position as a worldwide phenomenon.

*Bridgerton* also dominated international screens with 49 weeks in the Top 10 and a peak of 252 million hours.

Non-English hit *Money Heist* spent 33 weeks in the global Top 10, reaching 202 million peak hours, proving the power of Netflix’s international storytelling.

These global blockbusters highlight Netflix’s unique ability to deliver content that resonates across languages and cultures. By leveraging data-driven promotion, local-language support, and strategic marketing, Netflix ensures that its original content can capture global audiences simultaneously. Netflix continues to invest in content with global potential, ensuring that future releases have the opportunity to replicate the international success of these blockbusters.
